HLoRA: Hierarchical Bayesian Low-Rank Adaption for Multi-task Large Language Models

### 2024-06-06 paper/main.tex
rename hlora to bora

\usepackage[
backend=biber,
style=alphabetic,
sorting=ynt
]{biblatex}
\usepackage{amsmath}
\usepackage{amssymb}
\newcommand{\R}{\mathbb{R}}
\usepackage{booktabs}
\addbibresource{My Library.bib}

@@

\maketitle
\abstract{This paper introduces Bayesian Hierarchical Low-Rank Adaption (BH-LoRA), a novel method for finetuning multi-task Large Language Models (LLMs). Current finetuning approaches, such as Low-Rank Adaption (LoRA), perform exeptionally well in reducing training parameters and memory usage but face limitations when applied to multiple similar tasks. Practitioners usually have to choose between training separate models for each task or a single model for all tasks, both of which come with trade-offs in specialization and data utilization.

HLoRA addresses these trade-offs by leveraging a Bayesian hierarchical model that allows tasks to share information through global hierarchical prior parameters. This enables tasks with limited data to benefit from the overall structure derived from related tasks while allowing tasks with more data to specialize. We demonstrate the effectiveness of HLoRA on the next token generation task using the Talk of Norway Dataset, where each parliament representative's speeches represent distinct tasks. Our experimental results show that HLoRA outperforms both traditional individual and unified model approaches, achieving lower perplexity and better generalization across tasks. This method provides a scalable and efficient solution for multi-task LLM finetuning, with significant practical implications for diverse applications such as media content generation, chatbot responses, and scriptwriting.
}

@@

In this paper we suggest a way to circumvent these trade-offs and presents a generalization of the two LoRA fine-tuning methods for multi-task problems, which we call Hierarchical Low-Rank Adaption method (HLoRA).
The Hierarchical LoRA method allows tasks to share information between each other through sharing a set of global hierarchical prior parameters, in a Bayesian hierarchical setting. 
This means that tasks with little data can benefit from the global structure of the related tasks, while tasks with more data can specialize on their specific task.
We show in Section \ref{sec:hlora} that HLoRA can be seen as a generalization of the two approaches above.

To test our method, we evaluate the Hierarchical LoRA method on the next token generation problem of parliament speeches letting each parliament representative be a task. In our setting, HLoRA is superior to both conventional approaches above and suggest a way to mitigate the trade-off between the two.
% Write something about how easy it is to implement and how it can be used in practice?

The paper is organized as follows: Section \ref{sec:related_works} presents related works, Section \ref{sec:method} presents the Hierarchical LoRA method, (...)

@@

\subsection{The Hierarchical LoRA model} \label{sec:hlora}
In the hierarchical LoRA model, we introduce a LoRA parameter set for each task $d$, denoted as $\theta_d$, containing all the low-rank decomposition matricies $A_d$ and $B_d$ from each decomposed layer in the Large Language Model. The likelihood of the data $\mathcal{D}$ given the task parameters $\theta_{1:D}$ is then given by (combining equation \ref{eq:data} and \ref{eq:LLMprob}):
\begin{equation} \label{eq:lik}
    \text{L}(\mathcal{D} | \theta_1, \theta_2, \ldots, \theta_D) = \prod_{d=1}^D L(\mathcal{D}_d | \theta_d) = \prod_{d=1}^D \prod_{n=1}^{N_d} \prod_{i=1}^{W_n-1} \text{LLM}(w_{d,n,i+1} | w_{d,n,1:i}; \theta_d).
\end{equation}
%
%
Tasks will share some knowledge with each other, and we model this by introducing a Gaussian hierarchical prior over the task parameters $\theta_{1:D}$, given by
\begin{equation} \label{eq:prior}
    P(\theta_{1:D} | \Theta, \tau ) = \prod_{d=1}^D N(\theta_d ; \Theta, \frac{1}{\tau} I)
\end{equation}
%
where $\Theta$ is a set of hierarchical mean parameters, $I$ is a unit diagonal matrix and $\tau \geq 0$ is a scalar precision hyperparameter controlling how similar the task parameters $\theta_d$ should be to the hierarchical mean parameters $\Theta$.
We let $\Theta$ have a uniform hyperprior (i.e. $P(\Theta) = 1$). The prior assumes that the task parameters share the common origin from the shared Gaussian prior, because the tasks are related: input data come from similar distributions and the output tasks has similar aspects. 
The precision hyperparameter $\tau$ controls how much the task parameters $\theta_d$ are a priori expected to be allowed to deviate from the hierarchical mean parameters $\Theta$. 
This is essentially a measure for how much structure and information the tasks share with each other, and should be set by the practitioner based on her's prior knowledge.

@@

The resulting HLoRA model balances each task based on its data size and similarity to the other tasks: If a task has few data points, the posterior in equation \ref{eq:posterior} will be dominated by the prior term and the task parameters will be close to the global hierarchical parameters and borrow information and structure from these. On the other hand, if a task has a lot of data points, the likelihood will dominate the posterior and the task parameters will be allowed to learn more about the specific task and rely less on the hierarchical parameters.

@@

\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Our results in Figure \ref{fig:results_plot} show that the hierarchical LoRA model can improve the test perplexity for all tasks compared to both training the models independently and training one model for all tasks. This confirms our hypothesis that HLoRA can be a useful method for multi-task problems, and that it can mitigate the trade-off between training each task separately and training one model for all tasks.
It is interesting to note the HLoRA also helps for the tasks with relatively more data.

@@

\section{Conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This paper presented a generalization of the LoRA fine-tuning method for multi-task problems, which we call the Hierarchical LoRA method. The hierarchical model allows tasks to share information between each other through a set of global hierarchical prior parameters, and can be seen as a generalization of the two conventional approaches for multi-task problems. We show that the Hierarchical LoRA model can improve the test perplexity for all tasks compared to both training the models independently and training one model for all tasks, and that it can mitigate the trade-off between the two approaches.
Future work includes investigating the effect that HLORA has on more tasks and datasets, investigating how the relative capacity of the global model affects the performance, and investigating the full posterior distribution of the model.

>>>>

\usepackage[backend=biber,style=alphabetic,sorting=ynt]{biblatex}
\addbibresource{My Library.bib}
\usepackage{amsmath}
\usepackage{amssymb}
\newcommand{\R}{\mathbb{R}}
\usepackage{booktabs}
\addbibresource{My Library.bib}

@@

\maketitle
\abstract{This paper introduces Bayesian Hierarchical Low-Rank Adaption (BoRA), a novel method for finetuning multi-task Large Language Models (LLMs). Current finetuning approaches, such as Low-Rank Adaption (LoRA), perform exeptionally well in reducing training parameters and memory usage but face limitations when applied to multiple similar tasks. Practitioners usually have to choose between training separate models for each task or a single model for all tasks, both of which come with trade-offs in specialization and data utilization.

BoRA addresses these trade-offs by leveraging a Bayesian hierarchical model that allows tasks to share information through global hierarchical prior parameters. This enables tasks with limited data to benefit from the overall structure derived from related tasks while allowing tasks with more data to specialize. We demonstrate the effectiveness of BoRA on the next token generation task using the Talk of Norway Dataset, where each parliament representative's speeches represent distinct tasks. Our experimental results show that BoRA outperforms both traditional individual and unified model approaches, achieving lower perplexity and better generalization across tasks. This method provides a scalable and efficient solution for multi-task LLM finetuning, with significant practical implications for diverse applications such as media content generation, chatbot responses, and scriptwriting.
}

@@

In this paper we suggest a way to circumvent these trade-offs and presents a generalization of the two LoRA fine-tuning methods for multi-task problems, which we call Hierarchical Bayesian Low-Rank Adaption method (BoRA).
The Hierarchical LoRA method allows tasks to share information between each other through sharing a set of global hierarchical prior parameters, in a Bayesian hierarchical setting. 
This means that tasks with little data can benefit from the global structure of the related tasks, while tasks with more data can specialize on their specific task.
We show in Section \ref{sec:BoRA} that BoRA can be seen as a generalization of the two approaches above.

To test our method, we evaluate the Hierarchical LoRA method on the next token generation problem of parliament speeches letting each parliament representative be a task. In our setting, BoRA is superior to both conventional approaches above and suggest a way to mitigate the trade-off between the two.
% Write something about how easy it is to implement and how it can be used in practice?

The paper is organized as follows: Section \ref{sec:related_works} presents related works, Section \ref{sec:method} presents the Hierarchical LoRA method, Section \ref{sec:experimental_setup} describes the experimental setup, Section \ref{sec:results} presents the results, and Section \ref{sec:discussion} discusses the results and the implications of the Hierarchical LoRA method.

@@

\subsection{The Hierarchical Bayesian LoRA model} \label{sec:BoRA}
In the hierarchical LoRA model, we introduce a LoRA parameter set for each task $d$, denoted as $\theta_d$, containing all the low-rank decomposition matricies $A_d$ and $B_d$ from each decomposed layer in the Large Language Model. The likelihood of the data $\mathcal{D}$ given the task parameters $\theta_{1:D}$ is then given by (combining equation \ref{eq:data} and \ref{eq:LLMprob}):
\begin{equation} \label{eq:lik}
    \text{L}(\mathcal{D} | \theta_1, \theta_2, \ldots, \theta_D) = \prod_{d=1}^D L(\mathcal{D}_d | \theta_d) = \prod_{d=1}^D \prod_{n=1}^{N_d} \prod_{i=1}^{W_n-1} \text{LLM}(w_{d,n,i+1} | w_{d,n,1:i}; \theta_d).
\end{equation}
%
%
Tasks will share some knowledge with each other, and we model this by introducing a Gaussian hierarchical prior over the task parameters $\theta_{1:D}$, given by
\begin{equation} \label{eq:prior}
    P(\theta_{1:D} | \Theta, \tau ) = \prod_{d=1}^D N(\theta_d ; \Theta, \frac{1}{\tau} I)
\end{equation}
%
where $\Theta$ is a set of hierarchical mean parameters, $I$ is a unit diagonal matrix and $\tau \geq 0$ is a scalar precision hyperparameter controlling how similar the task parameters $\theta_d$ should be to the hierarchical mean parameters $\Theta$.
We let $\Theta$ have a uniform hyperprior (i.e. $P(\Theta) = 1$). The prior assumes that the task parameters share the common origin from the shared Gaussian prior, because the tasks are related: input data come from similar distributions and the output tasks has similar aspects. 
The precision hyperparameter $\tau$ controls how much the task parameters $\theta_d$ are a priori expected to be allowed to deviate from the hierarchical mean parameters $\Theta$. 
This is essentially a measure for how much structure and information the tasks share with each other, and should be set by the practitioner based on her's prior knowledge.

@@

The resulting BoRA model balances each task based on its data size and similarity to the other tasks: If a task has few data points, the posterior in equation \ref{eq:posterior} will be dominated by the prior term and the task parameters will be close to the global hierarchical parameters and borrow information and structure from these. On the other hand, if a task has a lot of data points, the likelihood will dominate the posterior and the task parameters will be allowed to learn more about the specific task and rely less on the hierarchical parameters.

@@

\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Our results in Figure \ref{fig:results_plot} show that the hierarchical LoRA model can improve the test perplexity for all tasks compared to both training the models independently and training one model for all tasks. This confirms our hypothesis that BoRA can be a useful method for multi-task problems, and that it can mitigate the trade-off between training each task separately and training one model for all tasks.
It is interesting to note the BoRA also helps for the tasks with relatively more data.

@@

\section{Conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This paper presented a generalization of the LoRA fine-tuning method for multi-task problems, which we call the Hierarchical LoRA method. The hierarchical model allows tasks to share information between each other through a set of global hierarchical prior parameters, and can be seen as a generalization of the two conventional approaches for multi-task problems. We show that the Hierarchical LoRA model can improve the test perplexity for all tasks compared to both training the models independently and training one model for all tasks, and that it can mitigate the trade-off between the two approaches.
Future work includes investigating the effect that BoRA has on more tasks and datasets, investigating how the relative capacity of the global model affects the performance, and investigating the full posterior distribution of the model.
